---
title: "Regression Analysis: Fuel Efficiency in Automobiles" 
author: "Sarah Agatoni"
output: html_document
---

<br> 

### Summary

This project uses data from the 1974 Motor Trend US magazine (comprising fuel consumption and 10 aspects of automobile design and performance for 32 automobiles) to:

- explore the relationship between various variables and fuel consumption

- compare the fuel efficiency of automatic versus manual transmission cars 

<br> 

**Environment**
```{r message=FALSE, warning=FALSE}
if (!require("pacman")) install.packages("pacman"); library(pacman)
pacman::p_load(dplyr, corrplot,  ggplot2, ggthemr, plotly)
```

<br>

### Data 

*load data*
```{r data}
data("mtcars")
```

<br>

*view first few rows*
```{r head}
head(mtcars, 4)
```
<br>

The data has 32 observations of the following variables:

- $mpg$: fuel consumption (miles/gallon)
- $cyl$: number of cylinders
- $disp$: displacement
- $hp$: gross horsepower
- $drat$: rear axle ratio
- $wt$: weight (1000 lbs)
- $qsec$: quarter mile time
- $vs$: engine shape (0 = v-shaped; 1 = manual)
- $am$: transmission (0 = automatic; 1 = manual)
- $gear$: number of forward gears 
- $carb$: number of carburetors 

<br> 

*check for missingness*
```{r na_check}
sum(is.na(mtcars))
```

<br>

*transforming numeric variables to factors*
```{r to_factor, message=FALSE}
mydata <- mtcars %>% mutate_at(c("cyl", "vs", "am", "gear", "carb"), as.factor)
```

<br> 


### Exploratory Analysis


<br>

**Fig 1: Correlation Plot**
```{r correlation}
corrplot.mixed(cor(mtcars), order="hclust", tl.col="black")
```

**Observations**: 

- Fuel consumption ($mpg$) is highly correlated with automobile weight ($wt$), horsepower ($hp$), number of cylinders ($cyl$) and displacement ($disp$). These are potential predictors.   

- These potential predictors are themselves correlated. The best model will not contain all of the potential predictors.

- Transmission type ($am$) is moderately correlated with the rear axel ratio ($drat$) and number of gears ($gear$). So, we might need to eliminate these variables to limit variance inflation due to collinearity.

<br> 


**Fig 2: Box plot: Fuel by Transmission**
```{r}
ggplot(mydata, aes(x=am, y=mpg, fill = am)) + 
  geom_boxplot() + 
  labs(x = "Type", y = "Miles per Gallon") +
  scale_fill_discrete(labels = c("Automatic", "Manual"))
```

**Observation**: Manual cars in the sample have better fuel efficiency i.e. travel farther on comparable fuel than automatic cars.

<br>

*how much farther?*
```{r}
auto <- mydata[mydata$am == 0,];man <- mydata[mydata$am == 1,]
mean(man$mpg) - mean(auto$mpg)
```

<br> 

*is this difference unique to our sample?*
```{r}
t.test(man$mpg, auto$mpg, alternative = c("two.sided"))
```

**Observations**:

- Everything else equal, manual cars *probably* have a fuel efficiency advantage over automatic cars

- However, 'everything else' need not be equal. As observed from $Fig. 1$, factors such as weight, horsepower and number of cylinders may better explain differences in fuel consumption. 

<br> 

**Fig 3: Box plot: Fuel by Transmission|Number of Cylinders **

```{r boxplot, cache=TRUE}
ggthemr("dust")

bp <- ggplot(mydata, aes(x=am, y=mpg, fill = am)) + 
    geom_boxplot() +
    facet_wrap(. ~ cyl, labeller = "label_both")

ggplotly(bp)
```

**Observation**: As the number of cylinder increases, the difference in fuel consumption between automatic and manual transmission cars shrinks.

<br>

### Modeling 

The goal is to gauge the effect of transmission type on fuel consumption after accounting for other predictors. And to investigate the best overall predictors of fuel consumption.

<br>
The model with only $am$ as a predictor is the baseline. 
```{r}
baseline <- lm(mpg ~ am, data = mydata)
summary(baseline)
```
**Observations**:

- In this model, transmission type is a significant predictor but it explains only 34% of the variance in fuel consumption

- If you knew nothing else about a car, you may *sometimes* guess correctly that a manual car will have better fuel efficiency than an automatic car, but you would almost certainly be wrong about the magnitude of the difference.

<br> 


**Variable Selection**

We use stepwise regression as a preliminary way of identifying important variables. 

```{r}
full <- lm(mpg ~., mydata)
select <- MASS::stepAIC(full, direction = "both", trace = FALSE)
```
The stepwise method identifies $cyl$, $hp$, $wt$ and $am$ as potential predictors. 

<br>

**Model Building**

```{r}
summary(select)$adj.r.squared
```

A model with these variables is much more informative (50% more explanatory power) but we still need to decide what combination of these predictors yields the best model.

<br>

To do so I fit a (partial) second order model and eliminate insignificant variables. This yields:
```{r}
chosen <- lm(mpg ~ hp + wt + I(hp^2) + I(wt^2), data = mydata)
summary(chosen)
```
This model has fairly high explanatory power (87.5%) with few, all-significant predictors. 

<br>

**NB** Adding transmission type does not improve the model's R-squared, in fact the model is penalized for having an extraneous variable.
```{r}
summary(lm(mpg ~ am + hp + wt + I(hp^2) + I(wt^2), data = mydata))$adj.r.squared
```
This implies that **knowing a car's weight and horsepower yields the most information about its fuel consumption**.

<br>

#### Diagnostics

Before proceeding, we check that the assumptions for linear regression aren't grossly violated. 

```{r}
par(mfrow=c(2,2)) 
plot(chosen)
par(mfrow=c(1,1))
```

<br>

*Zero Mean and Equal Variance of Errors* 

The `Residuals vs Fitted` shows residuals with a roughly zero mean zero and more or less equal spread, suggesting an unbiased model. 

<br>

*Normality of Errors* 

The points in the top-right plot seem to have deviations from normality (they do not lie on the diagonal line). We perform a secondary check for normality: 

```{r}
shapiro.test(summary(chosen)$resid)
```
The null hypothesis of this test is that the errors are normally distributed. With a p-value of 0.09, we fail to reject the null i.e. we find no evidence that normality is violated.

<br> 

*Influential Outliers*

The `Residuals vs Leverage` indicate the possibility of leverage points. To gauge influence, we look at the variables with Cook's Distance > 4/n

```{r}
which(cooks.distance(chosen) > 4/nrow(mydata))
```
However, these values do not appear to have been recorded in error and deleting them doesn't significantly improve the model. So we stick to the chosen model. 

<br>


### Conclusion 

```{r}
chosen
```

<br>

**Final Observations**:

- All things equal, a manual car will have better fuel efficiency than an automatic one. 

- Given the information we have, the best predictor of fuel efficiency is weight. 

- A car's horsepower, displacement and number of cylinders provide additional information. But seeing as they are all highly correlated, it is more or less the same information. Horsepower is just slightly more informative.

- Both weight and horsepower affect fuel efficiency in the opposite direction i.e. heavier cars and those with high horsepower consume more fuel.



__________________________________________________________END__________________________________________________________

